---
title: "DEP2 Analyze Demo"
author:
- name: Huanmu
output:
  BiocStyle::html_document
abstract: |
  This is a demo for goupmeeting.
vignette: |
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval = T,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

```{r include=FALSE}
library(BiocStyle)
library(dplyr)
library(tibble)
library(purrr)
library(SummarizedExperiment)
library(DEP2)
require(shiny)
require(shinydashboard)
require(shinyBS)
# library(shinydashboardPlus)
library(shinyvalidate)
library(shinyWidgets)
library(bslib)
library(ComplexHeatmap)
library(ggplot2)
library(dplyr)
library(tidyr)
library(limma)
library(colourpicker)
library(QFeatures)
library(grid)
library(fdrtool)
```

# Begin

DEP2 provide two kinds of proteomics analyze:

1.  Driectly from proteingroups- or enriched-peptides-level quantification

    Require proteomics or modified peptide table, containing a assay in wide format.

2.  Resummarize protein-level from peptide-level quantification

    Require overall peptide table, containing a assay in wide format.

## Directly from quantification
Require proteingroup or modified peptide table and experiment design (optional)

### Begin from proteingroups

```{r}
## read in PGs table
pg_table <- read.csv("example/proteinGroups(2).txt", sep = "\t")

## creat row identifiers name and ID using giving 'names' and 'ids' columns
pg_unique <- make_unique(pg_table, names = "Gene.names", ids = "Protein.IDs", delim = ";")
head(pg_unique[, c("name", "ID")])

## construct a SummarizedExperiment object
expression_cols <- grep("^LFQ.intensity", colnames(pg_unique),value = F)
PG_SE <- make_se_parse(proteins_unique = pg_unique, columns = expression_cols, mode = "delim")



## abundance assay
assay(PG_SE[1:5, 1:9])

## row infomations
rowData(PG_SE)[1:5, 1:4]

## col design
colData(PG_SE)
PG_SE <- make_se(proteins_unique = pg_unique, columns = expression_cols, 
                  expdesign = data.frame(colData(PG_SE))
                  )
```


### Begin from modified peptide

```{r}
## read in modified sites table
PTM_table <- read.csv("example/GlyGly (K)Sites.txt",sep = "\t")
## creat row identifiers name and ID using giving 'names' and 'ids' columns, copy "gene_name", "protein_ID"
PTM_unique <- make_unique_ptm(PTM_table, 
                             gene_name = "Gene.names", protein_ID = "Protein", 
                             aa = "Amino.acid", pos = "Position")

PTM_unique[1:5, c("name", "ID","gene_name","protein_ID")]

## construct a SummarizedExperiment object
grep("^(Intensity.)", colnames(PTM_unique),value = T)
grep("^(Intensity.)[^_]*_[0-9]+$", colnames(PTM_unique),value = T)
PTM_SE <- make_se_parse(PTM_unique, columns = grep("^(Intensity.)[^_]*_[0-9]+$", colnames(PTM_unique)), mode = "delim")
assay(PTM_SE)[1:5,]
colData(PTM_SE)
```


## Re-aggregation from peptide quantification
Read in a peptide table and creat a QFeatures object
```{r}
Pep_table <- read.csv("example/USP15KO_mef_peptides.txt",sep = "\t")
grep("^(Intensity.)", colnames(Pep_table),value = T) ## expression cols
head(Pep_table$Proteins) ## protein IDs

## construct QFeatures object
PEP_pe <- make_pe_parse(Peptide = Pep_table, ecols = grep("^(Intensity.)", colnames(Pep_table)), protein_col = "Proteins",mode = "delim")
class(PEP_pe)
experiments(PEP_pe)
class(PEP_pe[["peptideRaw"]])
colData(PEP_pe)
```

# Differentail analyze for proteomics


## Direct
*Filter*
```{r}
PG_filter <- filter_se(PG_SE, thr = 0, missnum = 2, filter_formula = ~Potential.contaminant!="+"& Reverse != "+")

nrow(PG_SE)
nrow(PG_filter)
plot_missval(PG_SE)
plot_missval(PG_filter)

PTM_filter <- filter_se(PTM_SE, thr = 0, missnum = NULL, filter_formula = ~Potential.contaminant!="+"& Reverse != "+")
```

*Normalization*
```{r}
PG_norm <- normalize_vsn(PG_filter)
plot_normalization(PG_filter, PG_norm)
plot_cvs(PG_filter)
plot_cvs(PG_norm) ##Plot sample coefficient of variation

PTM_norm <- normalize_vsn(PTM_filter)
```

*imputation*
```{r}
PG_imp <- DEP2::impute(PG_norm, fun = "MinDet")
plot_imputation(PG_norm, PG_imp)

PTM_imp <- DEP2::impute(PTM_norm, fun = "MinDet")
plot_imputation(PTM_norm, PTM_imp)
# PTM_imp <- impute(PTM_norm, fun = "RF")

```

*hypothesis testing*
```{r}
cat("weg")
PG_diff <- DEP2::test_diff(PG_imp, type = "control",control = "Biotin")
PG_res <- DEP2::add_rejections(PG_diff, thresholdmethod = "intersect" , alpha = 0.01, lfc  = 2)

PTM_diff <- DEP2::test_diff(PTM_imp, type = "control",control = "GFP", fdr.type = "Storey's qvalue") ## adjusted by Qvalue
PTM_res <- DEP2::add_rejections(PTM_diff, thresholdmethod = "curve" , curvature  = 1, x0_fold = 2)

DEP2::plot_heatmap(PG_res)
DEP2::plot_heatmap(PG_res, kmeans = T, k = 4)
DEP2::plot_heatmap(PG_res, kmeans = T, k = 4, split_order = 4:1) ## reorder clusters

(choose_pro <- DEP2:::get_signicant(PG_res) %>% rownames() %>% sample(.,10))
DEP2::plot_heatmap(PG_res, chooseToshow = choose_pro) ## choose some proteins
(DEP2:::get_contract(PG_res))
DEP2::plot_heatmap(PG_res, manual_contrast = "H2A_vs_Biotin") ## reduce contrast
DEP2::plot_ma_pro(PG_res,contrast = "H2A_vs_Biotin") ## MV proteins

```

## Re-aggregation

*filter*
```{r}
PEP_filter <- filter_pe(PEP_pe, thr = 1, missnum = 2, filter_formula = ~Potential.contaminant!="+"& Reverse != "+")
```

*normalize*
```{r}
# PEP_filter[["peptideImp"]] <- impute(PEP_filter[["peptideRaw"]], fun = "knn")

PEP_QF <- addAssay(PEP_filter, 
                   logTransform(PEP_filter[[1]]),
                   name = "peptidesLog")
PEP_QF <- QFeatures::normalize(PEP_QF, 
                    i = "peptidesLog",
                    name = "peptidesNorm",
                    method = "center.median")

PEP_QF <- addAssay(PEP_QF, 
                   DEP2::impute(PEP_QF[["peptidesNorm"]], fun = "knn"),
                   name = "peptidesImp")
experiments(PEP_QF)
plot_imputation(PEP_QF[["peptidesNorm"]], PEP_QF[["peptidesImp"]])
limma::plotDensities(assay(PEP_QF[["peptidesNorm"]]))
limma::plotDensities(assay(PEP_QF[["peptidesImp"]]))

```


```{r eval=T, include=TRUE}
# rowData(PEP_QF[["peptidesImp"]]) ->temp
# str(temp)
PEP_QF <- suppressWarnings(aggregate_pe(PEP_QF,
                                        aggrefun = "RobustSummary", aggregate_Peptide_Type = "Unique + Razor",
                                        fcol = "Proteins", peptide_assay_name = "peptidesImp", reserve = "Gene.names"
                          ))

experiments(PEP_QF)

## take the protein SummarizedExperiment out
pep_se = PEP_QF[["protein"]]
colnames(rowData(pep_se))
rowData(pep_se) <- make_unique(rowData(pep_se) %>% as.data.frame(), 
                                     names =  "Gene.names", 
                                     ids = "smallestProteingroups", 
                                     delim = ";")
rownames(pep_se) = rowData(pep_se)$name
pep_norm <- normalize_vsn(pep_se)

pep_diff = DEP2::test_diff(pep_norm, type = "control", control = "WT",fdr.type = "BH")
pep_res = add_rejections(pep_diff)

plot_heatmap(pep_res)
```



