---
title: "DEP2 Analyze Demo"
author:
- name: Huanmu
output:
  BiocStyle::html_document
abstract: |
  This is a demo for goupmeeting.
vignette: |
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval = T,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```


```{r include=FALSE}
library(BiocStyle)
library(dplyr)
library(tibble)
library(purrr)
library(SummarizedExperiment)
# library(DEP2)
library(DEP2)
require(shiny)
require(shinydashboard)
require(shinyBS)
# library(shinydashboardPlus)
library(shinyvalidate)
library(shinyWidgets)
library(bslib)
library(ComplexHeatmap)
library(ggplot2)
library(dplyr)
library(tidyr)
library(limma)
library(colourpicker)
library(QFeatures)
library(grid)
library(fdrtool)
library(magrittr)
library(clusterProfiler)
source("./R/functions_enrichment.R")

Cnetplot <- function (x, showCategory = 5, foldChange = NULL, layout = "kk",
                      colorEdge = FALSE, circular = FALSE, node_label = "all",
                      ...) {
  node_label <- match.arg(node_label, c("category", "gene",
                                        "all", "none"))
  if (circular) {
    layout <- "linear"
    geom_edge <- geom_edge_arc
  }
  else {
    geom_edge <- geom_edge_link
  }
  geneSets <- extract_geneSets(x, showCategory)
  g <- list2graph(geneSets)
  foldChange <- fc_readable(x, foldChange)
  size <- sapply(geneSets, length)
  V(g)$size <- min(size)/2
  n <- length(geneSets)
  V(g)$size[1:n] <- size
  if (colorEdge) {
    E(g)$category <- rep(names(geneSets), sapply(geneSets,
                                                 length))
    edge_layer <- geom_edge(aes_(color = ~category), alpha = 0.8)
  }
  else {
    edge_layer <- geom_edge(alpha = 0.8, colour = "darkgrey")
  }
  if (!is.null(foldChange)) {
    fc <- foldChange[V(g)$name[(n + 1):length(V(g))]]
    V(g)$color <- NA
    V(g)$color[(n + 1):length(V(g))] <- fc
    palette <- fc_palette(fc)
    p <- ggraph(g, layout = layout, circular = circular) +
      edge_layer + geom_node_point(aes_(color = ~as.numeric(as.character(color)),
                                        size = ~size)) + scale_color_gradientn(name = "log2 fold change",
                                                                               colors = palette, na.value = "#E5C494")
  }
  else {
    V(g)$color <- "#B3B3B3"
    V(g)$color[1:n] <- "#E5C494"
    p <- ggraph(g, layout = layout, circular = circular) +
      edge_layer + geom_node_point(aes_(color = ~I(color),
                                        size = ~size))
  }
  p <- p + scale_size(range = c(3, 10), breaks = unique(round(seq(min(size),
                                                                  max(size), length.out = 4)))) + theme_void()
  if (node_label == "category") {
    p <- p + geom_node_text(aes_(label = ~name), data = p$data[1:n,
    ], repel = TRUE)
  }
  else if (node_label == "gene") {
    p <- p + geom_node_text(aes_(label = ~name), data = p$data[-c(1:n),
    ], repel = TRUE)
  }
  else if (node_label == "all") {
    p <- p +
      geom_node_text(aes_(label = ~name), data = p$data, repel = TRUE, max.overlaps = Inf)
    # geom_node_text(aes_(label = ~name), data = p$data[-c(1:n),
    # ], repel = TRUE) +
    # geom_node_text(aes_(label = ~name), data = p$data[1:n,
    # ], repel = TRUE)
  }
  return(p)
}
```

# Begin

DEP2 provide two kinds of proteomics analyze:

1.  Driectly from proteingroups- or enriched-peptides-level quantification

    Require proteomics or modified peptide table, containing a assay in wide format.

2.  Resummarize protein-level from peptide-level quantification

    Require overall peptide table, containing a assay in wide format.

## Directly from quantification
Require proteingroup or modified peptide table and experiment design (optional)

### Begin from proteingroups

```{r}
## read in PGs table
pg_table <- read.csv("example/proteinGroups(2).txt", sep = "\t")

## creat row identifiers name and ID using giving 'names' and 'ids' columns
pg_unique <- make_unique(pg_table, names = "Gene.names", ids = "Protein.IDs", delim = ";")
head(pg_unique[, c("name", "ID")])

## construct a SummarizedExperiment object
expression_cols <- grep("^LFQ.intensity", colnames(pg_unique),value = F)
PG_SE <- make_se_parse(proteins_unique = pg_unique, columns = expression_cols, mode = "delim")

PG_filter <- filter_se(PG_SE, thr = 0, missnum = 2, filter_formula = ~Potential.contaminant!="+"& Reverse != "+")

PG_norm <- normalize_vsn(PG_filter)
PG_imp <- DEP2::impute(PG_norm, fun = "MinDet")

PG_diff <- DEP2::test_diff(PG_imp, type = "control",control = "Biotin")
PG_res <- DEP2::add_rejections(PG_diff, thresholdmethod = "intersect" , alpha = 0.01, lfc  = 2)

class(PG_res)
```

## enrichment analysis
```{r}
get_contrast = DEP2:::get_contrast
get_signicant = DEP2:::get_signicant
ORA_res = test_ORA(PG_res, type = "GO")
class(ORA_res)
# exctract significant terms
ORA_res_sig = get_ORA_result(ORA_res,pvalueCutoff = 0.05, qvalueCutoff = 0.2,return_table = F)
class(ORA_res_sig)
dotplot(ORA_res_sig)
barplot(ORA_res_sig)
# Cnetplot(x = ORA_res_sig, showCategory = 15, foldChange = NULL, circular = T, colorEdge = T, ont = "BP")

# exctract tables of significant terms by return_table = T
ORA_res_sig_table = get_ORA_result(ORA_res,pvalueCutoff = 0.05, qvalueCutoff = 0.2, return_table = T)
DT::datatable(head(ORA_res_sig_table))


```

```{r}
ORA_cluster_res = test_ORA(PG_res, type = "GO", by_contrast = T)
ORA_cluster_res_sig = get_ORA_result(ORA_cluster_res,pvalueCutoff = 0.05, qvalueCutoff = 0.2,return_table = F)
dotplot(ORA_cluster_res_sig)
```









